---
title: "Project Report II"
author: "Yoni Ackerman, Dario Cantore, Zhenqi Wang"
output:  ioslides_presentation
bibliography: papers/library.bib
---

```{r, echo = FALSE}
nc_files <- list.files(paste0("/accounts/grad/yoni/Documents/Stat222/",
                              "data/cmip5-ng/"),
                       recursive = TRUE, full.names = TRUE)
nc <- nc_files[[1]]
dat <- nc_open(nc)
lat <- ncvar_get(dat, "lat")
lon <- ncvar_get(dat, "lon")
dat <- ncvar_get(dat, "tas")
```

## Recap

Data is spatial timeseries output from:

- 46 climate models
- 36 Variables (surface temperature, shortwave radiation, rainfall, etc.)
- 10 Scenarios (historical starting values, different GHG projections, etc.)
- 47 Ensembles (averages of model runs - make predictions more robust)

Challenges:

- 17 Terabytes of high-dimensional data
- Model non-independence 
- Very little domain knowledge

## Quick Glimpse

```{r, echo = FALSE, fig.width = 8, fig.height = 5}

tmp <- dat[ , , 400]
tmp <- ldply(1:dim(tmp)[1], function(i){
    ldply(1:dim(tmp)[2], function(j){
       data.frame(lon = i, lat = j, tas = tmp[i, j])
       })
    })     

ggplot(tmp, aes(x = lon, y = lat, fill = tas)) +
	    geom_raster(interpolate = TRUE) +
            ggtitle("Variation in Surface Temp (K) at 400 months")
```

## Motivation
- How to build consensus model [@Knutti2010]?
- How non-independent are the models [@Hong 1996]?
- Do the independence relationships change through time and across scenarios?
- How can we best visualize these data and our results?

## Compression

## Correlations

## References